A Survey of Quantitative Descriptions of Molecular Structure

### 2012-07-22 paper.tex
Made some more edits to address review

Geometric descriptors require a 3D conformation as input and therefore
the computations are not as fast as topological descriptors which do
not require the 3D coordinate generation step. Nearly all geometric
descriptors work on a single conformation and for cases where multiple
conformation must be considered, an averaging procedure is usually
employed. Examples of these descriptors include the gravitational
index \cite{Katritzky:1996ly} and moment of inertia descriptors.
Geometric descriptors encompass many aspects of a molecular
structure. For example, a number of them characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors \cite{Stanton:1990aa} that characterize the
partial charge distribution over the molecular surface. Note that
surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tessellation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parameterized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups.

It is important to note that there is a difference between the
information a descriptor captures and how it is calculated. The above
categories are based on input requirements. This gives rise to
situations where the molecular volume property can be
described by geometrical descriptors, e.g. using an analytical
approaches ~\cite{Eisenhaber:1995qf}, and with approximate,
constitutional descriptors, such as the very fast algorithm use by the
VABC approach~\cite{Zhao:2003fk} which is based on group
contributions.  The latter descriptor has been implemented in the CDK
and correlates well with other approximations to the molecular volume
(Figure \ref{fig:molvol}).

@@

The last category consists of QM descriptors. These
descriptors can also be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening
applications). These descriptors can be evaluated at various levels of
theory. For example, partial charges, HOMO and LUMO energies are
commonly evaluated using semi-empirical methods, while Eroglu and
T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
such as electronegativity, hardness and ionization potential (though
Puzyn et al question the need for such a high level of theory
\cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk} employed
frontier orbital theory and DFT to generate quantum mechanical analogs
of Hammett substituent constants \cite{Hammett:1937aa} and showed that
they agreed well with empirically derived constants.

@@

\textbf{Descriptors for polymeric systems} These systems represent a
challenge for descriptor development. There have been reports that
employed traditional small-molecule descriptors
\cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
terms of the individual monomers. A number of approaches have been
described that includes both deterministic, geometry derived
descriptors \cite{Edvinsson:2003bh} as well as stochastic methods
\cite{Gonzales-Diaz:2003ly}. For the case of polypeptides the
descriptors generally tend to be based on the individual amino
acids. Examples include the VHSE family of descriptors described by
Mei et al \cite{Mei:2005kx}, a set of side chain descriptors from
Collantes and Dunn \cite{Collantes:1995vn}, field based descriptors
\cite{Norinder:1991ys} and the AAindex database
\cite{Kawashima:1999zr} which allows rapid lookup of amino acid
properties.

@@

\textbf{Bioactivity as a descriptor} While the descriptors mentioned
so far are derived from some form of chemical structure, an
alternative approach is to employ observed biological activities of a
molecule as descriptors of that molecule.  This approach was taken by
Sedykh et al \cite{Sedykh:2011fk} who hypothesized that dose response
data points from high throughput dose-response assays could be employed
as biological descriptors. They observed that when such descriptors
were combined with traditional chemical structure descriptors, the
predictive performance of models developed using the combined set was
better than those developed using conventional approaches. A similar
approach can be taken using the PASS methodology
\cite{Poroikov:2007aa}, though this is an indirect method - one must
predict the PASS profile from the chemical structure, which can then
be used in subsequent predictive models.

@@

While library calls for descriptor evaluation is the most flexible
approach, it is not useful for non-programmers as well as users who
simply need a descriptor matrix for modeling or searching. As a
result, GUI or command line wrappers around descriptor APIs enable
easier usage. The CDKDescUI application
(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
application that exposes the Chemistry Development Kit (CDK) descriptor and fingerprint API's. It
allows the users to specify a SMILES or SD file and evaluate a
selected set of descriptors or fingerprints and export them to a text
file. The application also has a command line mode that can be run in
server environments. A similar application is PaDEL \cite{Yap:2011fk}
that includes the CDK descriptors and also implements a number of
additional descriptors such as the extended topochemical atom
descriptors \cite{Roy:2004uq}, McGowan Volume \cite{Zhao:2003kx}, and
the Klekota \& Roth substructure counts \cite{Klekota:2008vn}.

@@

A topic closely related to versioning is descriptor naming. The
preceding sections have discussed a categorization of descriptors and
have used a number of commonly used names. When performing
calculations in an automated manner, or employing remote services for
descriptor calculations \cite{Dong:2007aa}, it is useful to be able to
refer to descriptors via a standardized naming scheme. Currently the
QSAR modeling community lacks such a scheme, though the CDK, a Java library for cheminformatics
\cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
specification'' approach defines a set of classes that include
information of the implementation title, identifier and vendor. The
identifier may or may not include a version number. The actual details
of the a descriptor are listed in a dictionary, along with an optional
namespace. The dictionary entry is referenced by the descriptor
specification and this design thus allows one to seamlessly work with
multiple implementations of the same descriptor from different vendors
and keep track of which version is employed throughout a calculation
or workflow. In addition, the CDK provides a hierarchical annotation
of descriptors, grouping them into different classes. Since this
annotation is based on RDF \cite{Taylor:2006ab}, it allows one to go
beyond simple groupings and perform reasoning on descriptors. A simple
example is to identify functionally equivalent descriptors (such as
simple transformations of a base descriptor - square and cube roots of
the gravitational index is an example \cite{Wessel:1998ve}) in an
automated fashion. While this infrastructure is currently only
supported by the CDK, adoption of it (or some similar framework) would
enable easier interoperability between descriptor implementations and
enhance reproducibility. The framework is, however, actively used by
software using the CDK descriptors, such as
Bioclipse~\cite{Spjuth2010} and Ambit~\cite{Jelizakova2011}. Moreover,
the ideas have been adopted by the CHEMINF ontology for
cheminformatics~\cite{Hastings2011}, which is used by the CHESS
framework which also uses the CDK for descriptor calculations~\cite{Chepelev2011b}.

@@

Molecular descriptors play a fundamental role in cheminformatic and
chemometric analyses and as we have described in this article it is
possible to evaluate thousands of descriptors using a variety of
software. Though the bulk of descriptors are well defined in the
literature, multiple implementations of the same descriptor can yield
different results. These differences arise from differences in the
underlying chemistry models and reference data used by the
implementations. As shown by Gupta et al \cite{Gupta:2010uq},
the performance of predictive models developed using commercial or
Open Source descriptor implementations is very similar. While
different tools differ in the specific number and types of descriptors
that they calculate, the fact many descriptors are correlated with
others suggests that it does not matter too much which set of
descriptors are used in an application. Of course, this cannot be a
general rule - specific applications may require a specific set of
descriptors. This is not to say that all descriptor tools exhibit
similar performance. For example, it is clear from Figure \ref{fig:logp}
that the ACD Labs implementation of log P performs significantly
better than other implementations. Though this is not a completely
rigorous comparison (the methods underlying the ACD Labs, CDK and
ChemAxon implementations are different), it does highlight that
certain implementations of a descriptor can fare better than others -
especially those cases where the descriptor is based on a predictive
model.

>>>>

Geometric descriptors require a 3D conformation as input and therefore
the computations are not as fast as topological descriptors which do
not require the 3D coordinate generation step. Nearly all geometric
descriptors work on a single conformation and for cases where multiple
conformation must be considered, an averaging procedure is usually
employed. Examples of these descriptors include the gravitational
index \cite{Katritzky:1996ly} and moment of inertia descriptors.
Geometric descriptors encompass many aspects of a molecular
structure. For example, a number of them characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors (CPSA) \cite{Stanton:1990aa} that
characterize the partial charge distribution over the molecular
surface. Note that surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tessellation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parameterized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and MOE van der Waals Surface Area (VSA) descriptors
\cite{Labute:2008aa} that are based on precalculated surface area
values derived from a list of functional groups.

It is important to note that there is a difference between the
information a descriptor captures and how it is calculated. The above
categories are based on input requirements. This gives rise to
situations where the molecular volume property can be described by
geometrical descriptors, e.g. using an analytical approaches
~\cite{Eisenhaber:1995qf}, and with approximate, constitutional
descriptors, such as the very fast algorithm use by the van der Waals
Volume as a Sum of Atomic and Bond Contributions (VABC)
approach~\cite{Zhao:2003fk} which is based on group contributions.
The latter descriptor has been implemented in the Chemistry
Development Kit (CDK) and correlates well with other approximations to
the molecular volume (Figure \ref{fig:molvol}).

@@

The last category consists of QM descriptors. These descriptors can
also be applied to problems such as Quantitative Structure-Activity
Relationship (QSAR) models (though of course the
time complexity of QM derived descriptors do not make them amenable
for large scale virtual screening applications). These descriptors can
be evaluated at various levels of theory. For example, partial
charges, Highest Occupied Molecular Orbital (HOMO) and Lowest Occupied
MOlecular Orbital (LUMO) energies are commonly evaluated using
semi-empirical methods, while Eroglu and T\"{u}rkmen
\cite{Eroglu:2007uq} described density functional theory (DFT) derived
descriptors such as electronegativity, hardness and ionization
potential (though Puzyn et al question the need for such a high level
of theory \cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk}
employed frontier orbital theory and DFT to generate quantum
mechanical analogs of Hammett substituent constants
\cite{Hammett:1937aa} and showed that they agreed well with
empirically derived constants.

@@

\textbf{Descriptors for polymeric systems} These systems represent a
challenge for descriptor development. There have been reports that
employed traditional small-molecule descriptors
\cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
terms of the individual monomers. A number of approaches have been
described that includes both deterministic, geometry derived
descriptors \cite{Edvinsson:2003bh} as well as stochastic methods
\cite{Gonzales-Diaz:2003ly}. For the case of polypeptides the
descriptors generally tend to be based on the individual amino
acids. Examples include the Vector of Hydrophobic, Steric, and
Electronic (VHSE) family of descriptors described by Mei et al
\cite{Mei:2005kx}, a set of side chain descriptors from Collantes and
Dunn \cite{Collantes:1995vn}, field based descriptors
\cite{Norinder:1991ys} and the AAindex database
\cite{Kawashima:1999zr} which allows rapid lookup of amino acid
properties.

@@

\textbf{Bioactivity as a descriptor} While the descriptors mentioned
so far are derived from some form of chemical structure, an
alternative approach is to employ observed biological activities of a
molecule as descriptors of that molecule.  This approach was taken by
Sedykh et al \cite{Sedykh:2011fk} who hypothesized that dose response
data points from high throughput dose-response assays could be
employed as biological descriptors. They observed that when such
descriptors were combined with traditional chemical structure
descriptors, the predictive performance of models developed using the
combined set was better than those developed using conventional
approaches. A similar approach can be taken using the Prediction of
Biological Activity Spectra for Chemical Substances (PASS) methodology
\cite{Poroikov:2007aa}, though this is an indirect method - one must
predict the PASS profile from the chemical structure, which can then
be used in subsequent predictive models.

@@

While library calls for descriptor evaluation is the most flexible
approach, it is not useful for non-programmers as well as users who
simply need a descriptor matrix for modeling or searching. As a
result, GUI or command line wrappers around descriptor APIs enable
easier usage. The CDKDescUI application
(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
application that exposes the CDK descriptor and fingerprint API's. It
allows the users to specify a SMILES or SD file and evaluate a
selected set of descriptors or fingerprints and export them to a text
file. The application also has a command line mode that can be run in
server environments. A similar application is PaDEL \cite{Yap:2011fk}
that includes the CDK descriptors and also implements a number of
additional descriptors such as the extended topochemical atom
descriptors \cite{Roy:2004uq}, McGowan Volume \cite{Zhao:2003kx}, and
the Klekota \& Roth substructure counts \cite{Klekota:2008vn}.

@@

A topic closely related to versioning is descriptor naming. The
preceding sections have discussed a categorization of descriptors and
have used a number of commonly used names. When performing
calculations in an automated manner, or employing remote services for
descriptor calculations \cite{Dong:2007aa}, it is useful to be able to
refer to descriptors via a standardized naming scheme. Currently the
QSAR modeling community lacks such a scheme, though the CDK, a Java
library for cheminformatics \cite{Steinbeck:2006aa} does implement
such a scheme. The ``descriptor specification'' approach defines a set
of classes that include information of the implementation title,
identifier and vendor. The identifier may or may not include a version
number. The actual details of the a descriptor are listed in a
dictionary, along with an optional namespace. The dictionary entry is
referenced by the descriptor specification and this design thus allows
one to seamlessly work with multiple implementations of the same
descriptor from different vendors and keep track of which version is
employed throughout a calculation or workflow. In addition, the CDK
provides a hierarchical annotation of descriptors, grouping them into
different classes. Since this annotation is based on the Resource
Description Framework (RDF) \cite{Taylor:2006ab}, it allows one to go
beyond simple groupings and perform reasoning on descriptors. A simple
example is to identify functionally equivalent descriptors (such as
simple transformations of a base descriptor - square and cube roots of
the gravitational index is an example \cite{Wessel:1998ve}) in an
automated fashion. While this infrastructure is currently only
supported by the CDK, adoption of it (or some similar framework) would
enable easier interoperability between descriptor implementations and
enhance reproducibility. The framework is, however, actively used by
software using the CDK descriptors, such as
Bioclipse~\cite{Spjuth2010} and Ambit~\cite{Jelizakova2011}. Moreover,
the ideas have been adopted by the CHEMINF ontology for
cheminformatics~\cite{Hastings2011}, which is used by the CHESS
framework which also uses the CDK for descriptor
calculations~\cite{Chepelev2011b}.

@@

Molecular descriptors play a fundamental role in cheminformatic and
chemometric analyses and as we have described in this article it is
possible to evaluate thousands of descriptors using a variety of
software. Though the bulk of descriptors are well defined in the
literature, multiple implementations of the same descriptor can yield
different results. These differences arise from differences in the
underlying chemistry models and reference data used by the
implementations. As shown by Gupta et al \cite{Gupta:2010uq}, the
performance of predictive models developed using commercial or Open
Source descriptor implementations is very similar. While different
tools differ in the specific number and types of descriptors that they
calculate, the fact many descriptors are correlated with others
suggests that it does not matter too much which set of descriptors are
used in an application. Of course, this cannot be a general rule -
specific applications may require a specific set of descriptors. For
example, a predictive model for boiling points may not peform very
well if branching and intermolecular forces are not taken into account
by the descriptors in the model. This is not to say that all
descriptor tools exhibit similar performance. For example, it is clear
from Figure \ref{fig:logp} that the ACD Labs implementation of log P
performs significantly better than other implementations. Though this
is not a completely rigorous comparison (the methods underlying the
ACD Labs, CDK and ChemAxon implementations are different), it does
highlight that certain implementations of a descriptor can fare better
than others - especially those cases where the descriptor is based on
a predictive model.
